Flooding of the Okavango Delta influences Connectivity for Dispersing African Wild Dogs

### 2022-11-03 Chapter_8/04_Manuscript/Manuscript.tex
Various Updates
Updates addressing reviewer comments

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \orcid{0000-0003-3477-4365} \and
  Gabriele Cozzi\textsuperscript{1,2} \orcid{0000-0002-1744-1940} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \orcid{0000-0001-7477-2642} \and
  Dominik M. Behr\textsuperscript{1,2} \orcid{0000-0001-7378-8538}
}

@@

\textsuperscript{\S} Corresponding author (david.hofmann2@uzh.ch)

@@

\textbf{Running Title:} None

@@

\textbf{Keywords:} Seasonal Flooding of the Okavango Delta and its Consequences
for African Wild Dog Connectivity

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Climate change is expected to profoundly impact vital rates of wild living
animal populations. While the impact of climate change on the demographics of
local subpopulations has been studied repeatedly, little is known about the
consequences of environmental change on dispersal and connectivity between such
subpopulations.

Here, we capitalize on natural, flood-pulse driven environmental change across
the Okavando Deltas in northern Botswana to investigate the impact of changing
environmental conditions on the dispersal ability of the endangered African wild
dog. For this, we utilize a previously parametrized dispersal model and apply it
to simulate dispersal under two extreme environmental scenarios; one assuming
minimal flood extent, one assuming maximum flood extent.

In genral, our results suggest only a marginal impact of the flood extent on
dispersal prospects. Even during maximum flood, dispersal between neighboring
patches remained high. Between a few selected patches, however, dispersal was
substantially restricted during times of high flood. That is the Okavango delta
posed a substantial dispersal barrier that reduced connectivity between patches
during maximum flood, but revealed vital dispersal corridors that facilitated
dispersal into neighboring areas during minimum flood. Besides a better
understanding of the conservation needs for the African wild dog, our study also
presents a first step towards incorporating environmental change due to
seasonality or climate change. It also emphasizes the usefulness of
individual-based dispersal simulations as a conservation tool to study impacts
of environmental change in a spatially realistic way.

@@

\section{Introduction}
\subsection{Climate Change}

Despite the importance of climate change in determining species viability,
predicting the way and extent to which it will affect species remains
challenging. One exciting avenue to gain insights into the implications of
environmental change is to utilize seasonal variation in the environment.

@@

\subsection{Vital Rates}
Several studies investigating the impacts of climate change on life-history
traits have revealed that population persistence may critically depends on
climate conditions \citep{Paniw.2019}. While many studies examine climate
impacts on survival and reproductive rates, the influence of global change or
climatic extreme events on dispersal prospects has received little attention.

Ecosystems that undergoe seasonal environmental variation, in particular arid
regions that experience extreme climatic conditions, are among the most
vulnerable to climate change.

\subsection{Dispersal}
\subsection{Connectivity}
Dispersal is an important, if not the important, driver of landscape
connectivity and therefore of major interest to conservation authorities.

\subsection{Okavango Delta and African Wild Dogs}
The Okavango delta in Southern Africa poses a unique opportunity to study the
impacts of environmetal change on species dispersal ability and connectivity in
large scale natural experiment setup. According to projections by the IPCC,
Botswana and its surroundings are among the most vulnerable to climate change.

\section{Methods}
We conducted all analyses using the programming language R.

\subsection{Study Area}
The study area for this analysis was focused on the Okavango delta and its
surroundings in Southern Africa, comprising parts of Angola, Namibia, Botswana,
Zimbabwe, and Zambia (\Cref{StudyArea}). To accommodate for the long distance
dispersal events commonly observed in African wild dogs (e.g.
\citep{DaviesMostert.2012, Masenga.2016, Cozzi.2020}), we considered a large
rectangular extent stretching from 20\degree 30' E to 26\degree E. totaling to
an area of 300'000 km\textsuperscript{2}. The dominant geographical feature in
this region is the Okavango delta (OD), the world's largest inland delta and the
main driver of seasonal environmental change in the region. The flood-pulsing
rhythm of the OD is mainly dictated by precipitation in the Angolian highlands,
where rainfalls are channeled into the OD through the Okavango River. Although
precipitation in Angola peaks in ..., water only slowly descends through the
Okavango river and its tributaries, reaching the distal ends (i.e. the faults)
of the delta in July or August. At minimum extent, the flood covers an area of
3'600 km\textsuperscript{2}, whereas during maximum flood it covers more than
9'000 km\textsuperscript{2}. Vegetation in the study area is dominated by dense
mopane forest, mixed acacia woodland, and plain grassland. Human influence in
the study area is low and mainly concentrated around small villages at the
western periphery of the delta and to the city of Maun at the southern tip of
the delta. Large portions of land are dedicated national parks, game reserves or
forest reserves. The study area is also part of the world's largest
transboundary conservation initiative, the Kavango-Zambezi Transfrontier
Conservation Area.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_StudyArea.png}
  \caption{Study area across which we simulated dispersal events. Virtual
  dispersers were released at random locations within the orange source areas.}
  \label{StudyArea}
  \end{center}
\end{figure}

\subsection{Spatial Habitat Layers}
We represented the physical landscape through which dispersers could move by a
set of spatially referenced habitat layers each resolve at 250m x 250m. The set
of layers included water-cover, distance-to-water, tree-cover,
shrub/grassland-cover, and a human influence layer depicting anthropogenic
influences through villages, roads, and agriculture. A detailed description of
the different layers is provided in \Cref{Hofmann.2021}. Importantly, the
water-cover and the derived distance-to water layers were generated using MODIS
Terra satellite imagery, which enabled us to generate weekly updated layers that
provided detailed information about the flood-extent at any given point in time.
We had a total of 8xx remote sensed floodmaps at our disposal and used them to
generate two extreme scenarios; a minimal and maximum flood scenario. To create
the minimum flood scenario, we tallied the 50 floodmaps with smallest flood
extent into an average image. Finally, we created a binary map... Similarly, we
created an average image for high flood using the 50 most flooded maps. The
resulting maps are depicted in \Cref{FloodExtent}. For completeness we also
generated an  ``average'' floodmap by averaging across all 800 floodmaps
available to us. The results based upond this layer are presented in the
Appendix.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_FloodExtent.png}
  \caption{}
  \label{FloodExtent}
  \end{center}
\end{figure}

\subsection{Dispersal Model}
Our dispersal model was based on a previously parametrized and validated
\textit{integrated step-selection function (iSSF)} applied to the dispersal data
of 16 dispersing African wild dogs inhabiting the surroundings of the Moremi
National Park in the Okavango delta \citep{Hofmann.2021}. The iSSF model
consists of two complementary ``kernels'' plus their interactions. The first
kernel is a movement kernel and describes general movement behavior of
dispersing AWDs, irrespective of habitat conditions. The second kernel is a
habitat kernel and describes preferences of AWDs with regards to environmental
conditions. Finally, the model also includes interactions among the two kernels
and therefore allows to render how movement behavior changes depending on
habitat conditions.

\subsection{Source Areas}
We simulated dispersing AWDs originating from nine distinct source patches
located in the vicinity of the Okavango delta. The source areas were generated
as follows. First, we overlayed the OD with an oval that was bound by
geographical landmarks; in the north, the oval was bound by the Inflow of the
Okavango river into the ``pan-handle'' of the OD, north-east the oval was bound
by the Selinda-Spillway and the Linyanty swamp, towards South-East it was bound
by the Boteti river, and towards South-West by Lake Ngami. We then dissected the
polygon into five distinct patches using the same natural landmarks
(\Cref{StudyArea}). Patch one was given by the area south of the Boteti river
and area two by the area north of it up until the panhandle. Area three
stretched from the region easf of Maun towards north until the Selinda-spillway,
whereas area four stretched north of the spillway until west towards the
panhandle. Finally, a fifrth source area marked the peninsuale in the center of
the OD. In addition to these five source patches, we also distributed small
peripheral patches that we used to simulate individuals immigrating
\textit{into} system and to keep track of individuals leaving it.

\subsection{Dispersal Simulation}
For both environmental scenarios we simulated 1'000 individuals dispersing from
each of the nine source area depicted in \Cref{StudyArea}. The simulation
algorithm was based on the algorithm described in \Cref{Hofmann.2022} and works
as follows. A random location within the source area is chosen as a starting
point. Originating from the starting point, a set of 25 random steps are
generated by sampling step lengths from a gamma distribution fitted to observed
steps (what is a step?) (shape = , scale = ) and turning angles from a uniform
distibution on ($-\pi, +\pi$). Along each random step the underlying spatial
covariates are extracted and relevant movement metrics are computed (i.e.
log(sl), cos(ta), ta). The parametrized dispersal model is used to predict the
probability of each step for being chosen given the steps covariate values. One
of the steps is sampled based on assigned probabilities and the location of the
animal is updated. The procedure is then repated until the desired number of
steps (in our case 2000) is realized. Resulting trajectories resemble correlated
random walks. As the original model was trained using 4-hourly steps, a
simulated step also resembled straight line movements within four hours.

@@

\subsection{Connectivity}
Based on simulated dispersal trajectories in the two scenarios we quantified
connectivity in three complementary ways. First, we generated heatmaps depicting
the frequency at which different areas in the landscape were visited by
simulated dispersers. Such heatmaps serve to detect dispersal hotspots but are
less suitable to detect pinchpoints and bottlenecks. Hence, we also computed
spatially explicit betweenness scores which are useful to highlight such
pinchpoints (Bastille). To compute betweenness, we overlayed the study area with
a regular grid with 2.5 km x 2.5 km grid cells and determined how often
simulated individuals transitioned from one grid-cell to another. In cases where
the same individual repeatedly realized the same cell-transition, we only
counted a single transition to avoid emphasizing regions regions in which
individuals moved in circles. Finally, we calculated the number of successful
dispersal events between the different source areas. We coin this type of
connectivity ``inter-patch connectivity''. Dispersal between two areas was said
to be successful whenever a trajectory leaving one area intersected with the
target area. To gauge the dispersal duration needed to move between patches (be
consistent with ``patches'', ``source-areas'' etc.), we also recorded the
minimum number of steps that individuals had to move to move between areas.

\section{Results}
Heatmaps produced from simulated dispersal trajectories reveal that the OD acts
as a major dispersal barrier during periods of high flood, but reveals viable
dispersal corridors during periods of low flood. In particular, our analysis
reveals that the area north-west of Maun is frequently visited by dispersers at
in the minimum-flood scenario, yet that an extended flood, coupled with high
human influence around the city of Maun, enforces simulated individuals to
either detour around Maun, or to slip through the narrow window between village
and flood. Similarly, the betweenness map indicates the presence of a dispersal
corridor extending from the western section of the delta leading across the
delta's center into its western part, yet only during low-flood. During maximum
flood, the same corridor is pushed south and enforces dispersers to move closer
to the city of Maun. In general, connectivity during maxium flood is lower
compared to the minimum flood scenario (number links vs. number liks). Most
notably, source area 5 is almost inaccessible during maximum flood.

 Most notably, the
area north-west of Maun acts as dispersal habitat during phases of low flood,
yet is almost inaccessible during phases of maxium flood. This is reflected in
the heatmap, where connectivity across the western part of the delta is low
during maximum flood, but also in the betweenness map, where an important
movement corridors through the delta is only visible during minimum flood.
Although connectivity across large parts of the landscape remains unchanged
across environmental scenarios, one critical movement corridor shifts in
response to an expanding flood. Most notably, a corridor that links the western
and eastern part of the delta during minimum flood gets redirected further
south, thus leading more closely to the city of Maun. Unsurprisingly, the
differences are more pronounced for areas directly affected by the flood, yet it
is worth noting that even some areas not directly affected by the flood.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Heatmaps.png}
  \caption{}
  \label{Heatmaps}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_DifferenceMap.png}
  \caption{}
  \label{DifferenceHeatmaps}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Betweenness.png}
  \caption{}
  \label{Betweenness}
  \end{center}
\end{figure}

\input{IPC_Frequency.tex}
\input{IPC_Duration.tex}

@@

\cite{MurrayHudson.2006} predicted that increased temperatures, additional human
abstractions, and reduced river flows might lead to a ``Delta dying'' and that
the impact of climate will be much more pronounced than the impact of
anthropogenic water use.

@@

@@

\newpage
\begingroup
\singlespacing
\bibliography{Literatur}
\endgroup

>>>>

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \orcid{0000-0003-3477-4365} \and
  Dominik M. Behr\textsuperscript{1,2} \orcid{0000-0001-7378-8538} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \orcid{0000-0001-7477-2642} \and
  Gabriele Cozzi\textsuperscript{1,2} \orcid{0000-0002-1744-1940}
}

@@

\textsuperscript{\S} Corresponding author: david.hofmann2@uzh.ch

@@

\textbf{Running Title:} Seasonal Flooding of the Okavango Delta and its
Consequences for African Wild Dog Dispersal and Connectivity

@@

\textbf{Keywords:} movement, connectivity, seasonality, dispersal, conservation,
permeability

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Climate change is expected to profoundly impact the life history of wild-living
animal populations. While the impact of climate change on the demographics of
local subpopulations has been studied repeatedly, little is known about the
consequences of environmental change on dispersal and connectivity.

We capitalize on a ``natural experimental setup'', the flood-pulse driven
environmental change across the Okavango Delta in northern Botswana, to
investigate the impact of changing environmental conditions on dispersal
patterns and connectivity of the endangered African wild dog (\textit{Lycaon
pictus}). For this, we simulate dispersal trajectories across the landscape
under two extreme environmental scenarios; one assuming a maximum flood extent,
one assuming a minimum flood extent.

During maximum flood, we observe a reduction in connectivity and an increase in
dispersal durations between distinct habitat patches. Notably, dispersal into
the center of the Okavango delta is reduced by 80\% during maximum flooding. At
minimum flooding, conversely, the delta reveals vital dispersal corridors and
increases chances of successful dispersal into neighboring habitat patches.
Concomittantly, dispersal durations to move between patches is reduced at low
flood levels.

Climate change is expected to critically impact the flood-pulsing cycles of the
Okavango delta and may therefore critically alter landscape connectivity in this
area. Besides a better understanding of the conservation needs for the African
wild dog, our study also presents a first step towards incorporating
environmental changes due to seasonality or climate change in dispersal and
connectivity analyses. Finally, our analysis demonstrates the usefulness of
individual-based dispersal simulations as a pertinent conservation tool to study
impacts of environmental change while accounting for dispersal movements.

@@

\section{Introduction}
\subsection{Climate Change}
Climate change is expected to profoundly impact ecosystems worldwide, with
far-reaching consequences for species living therein (cite IPCC)
\citep{Ozgul.2010, Radchuk.2019}. However, empirical information on species' responses to climate change remains
scarce, especially for regions most vulnerable to altered conditions
\citep{Paniw.2021}.

\subsection{Dispersal}
The lack of knowledge in terms of species ability to cope with climate change is
further compounded by a lack of information on dispersal under changing
environmental conditions \citep{Travis.2013}. By bringing individuals away from
their natal location, dispersal strongly influences population dynamics
\citep{Clobert.2012}; it enables genetic exchange \citep{Frankham.2002,
Leigh.2012, Baguette.2013}, facilitates the colonization of empty habitats
\citep{Gustafson.1996, Hanski.1999b, MacArthur.2001}, and promotes reinforcement
of weakened and small subpopulations \citep{Brown.1977}. Dispersal also serves
as a means to track suitable habitat \citep{Raia.2012} and may thus serve as a
means to mediate the demographic consequences of environmental change
\citep{Hodgson.2009, Travis.2013}. Dispersal is also inextricably linked to the
concept of landscape connectivity, which is generally understood as the degree
by which the landscape facilitates or impedes movements among habitat patches
\citep{Tylor.1993}.

As previously highlighted by \cite{Travis.2013}, climate may affect dispersal
directly, by altering the propensity of individuals to disperse, or indirectly,
through changes in the biophysical environment. Here, we will focus on the
latter and study dispersal prospects under changing environmental conditions.

In spite of its importance for determining population dynamics and landscape
connectivity, our understanding of dispersal and how dispersal is affected by
changing environmental conditions is limited. This is owed to the difficulty of
obtaining data collected on dispersing animals at the appropriate temporal and
spatial scale \citep{Graves.2014, Vasudev.2015} and our inability to project
dispersal prospects under changing environmental conditions into the future.

The ability of species
to persist under these changing conditions depends on the amplitude and speed of
environmental change and species' responses to such changes (cite someone).

It is, in fact, presumed that humanity has already transgressed the planetary
boundary for climate change, therefore triggering irreversible feedbacks in the
biosphere \citep{Rockstrom.2009}.

@@

\subsection{Connectivity}
Dispersal is an important, if not the important, driver of landscape
connectivity and therefore of major interest to conservation authorities. It has
also been demonstrated that dispersal may lead to population declines in areas
where anthropogenic mortality is high and dispersal prospects low
\citep{Leigh.2012}.

\subsection{Okavango Delta and African Wild Dogs}
The Okavango delta in Southern Africa poses a unique opportunity to study the
impacts of environmetal change on species dispersal ability and connectivity in
large scale natural experiment setup. According to projections by the IPCC,
Botswana and its surroundings are among the most vulnerable to climate change.
While global temperatures are expected to increase by xx degrees until the end
of the 21st century, For example, \cite{Engelbrecht.2015} predicts an increase
between 4 and 6\degree C for regions in southern Africa. One species that
heavily relies on dispersal for long-term persistence and that inhabits such an
arid ecosystem in Southern Africa is the African wild dog (\textit{Lycaon
pictus}). The African wild dog is a lightly built canid lives in cohesive packs
comprising up to 20 individuals. Wild dogs are social breeders and the majority
of reproduction is monopolized by the dominant alpha couple who are supported by
subordinates in taking care of their pups. The species exhibits strong
inbreeding avoidance, hence juvenile individuals born into the pack disperse in
same-sex sibling coalitions after reaching sexual maturity \citep{McNutt.1996,
Behr.2020}. Notably, the African wild dog is the sole extant representative of
its genus and thus considered as high-priority species for conservation efforts
\citep{Leigh.2012}. However, fragmentation and destruction of their historic
habitat may have decreased dispersal success for this species and thus
jeopardizes connectivity \citep{Leigh.2012}. Since the impacts of climate change
are not uniformly distributed across the planet, but particularly pronounced at
high latitudes and arid ecosystems, the species may be seen as most vulnerable
to such ecosystem shifts.

\subsection{Our Study}
Here, we capitalize on a previously parametrized dispersal model and use
dispersal simulations to investigate connectivity patterns for African wild dogs
under two extreme scenarios; one assuming maximum flooding of the Okavango delta
and one assuming minimum flooding of the Okavango delta. Given that water poses
a barrier to dispersing individuals, we anticipate that dispersal prospects and
connectivity during maximum flood are low. During minimum flood, in contrast, we
expect to reveal the presence of several dispersal corridors that may be used to
freely and quickly move between remaining habitat patches.

\section{Materials and Methods}
We conducted all analyses using the programming language \textsf{R}. Any spatial
data manipulation was completed using the \textsf{terra} and \textsf{spatstat}
packages. Several helper functions for the simulation algorithm were written in
\textsf{C++} and imported to R using the \textsf{Rcpp} package. Figures were
generated using \textsf{ggplot2}, \textsf{igraph}, and \textsf{ggnetwork}. All
R-scripts used to conduct our analyses are provided through an online
repository.

\subsection{Study Area}
The study area for this analysis was focused on the Okavango delta (OD) and its
surroundings in Southern Africa, comprising parts of Angola, Namibia, Botswana,
Zimbabwe, and Zambia (\Cref{StudyArea}). The OD is the world's largest inland
delta and the main driver of seasonal environmental change in the region. While
our primary focus lied on the immediate surroundings of the Okavango Delta, we
considered a large rectangular extent stretching from 20\degree 30' E to
26\degree E. (totaling to an area of 300'000 km\textsuperscript{2}) to
accommodate for the long distance dispersal events commonly observed in African
wild dogs (e.g. \citealp{Davies-Mostert.2012, Masenga.2016, Cozzi.2020}). The
flood-pulsing rhythm of the OD is mainly dictated by precipitation in the
catchment areas of the Angolian highlands, where rainwater is collected and
channeled into the OD through the Okavango River. Although precipitation in
Angola peaks between December and March, water only slowly descends through the
Okavango river and its distributaries, reaching the distal ends of the delta in
July or August, where the water percolates at the Thamalakane and Kunyere
Faults. At minimum extent, the flood covers an area of 3'600
km\textsuperscript{2}, during maximum flood more than 9'000
km\textsuperscript{2}. Vegetation in the study area is dominated by mopane
forest, mixed acacia woodland, and grassland. Human influence is low and mainly
concentrated around small villages at the western periphery of the delta as well
as the city of Maun at the south-eastern tip of the OD. Large portions of land
are dedicated national parks, game reserves or forest reserves. The study area
is also part of the world's largest transboundary conservation initiative, the
Kavango-Zambezi Transfrontier Conservation Area, which aims to restore
connectivity between protected areas in Southern Africa.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_StudyArea.png}
  \caption{Study area across which we simulated dispersal. Simulated dispersers
  were released at random locations within the orange source areas distributed
  across the delta. Emigration zones (purple) served as checkpoints and enabled
  us to identify if and where simulated dispersers left the close surroundings
  of the Okavango delta. These zones were generated using a set of cutlines
  originating from the center of the delta and roughly cutting an elliptical
  buffer zone into sections of equal size.}
  \label{StudyArea}
  \end{center}
\end{figure}

\subsection{Spatial Habitat Layers}
We represented the physical landscape through which dispersers could move by a
set of spatially referenced habitat layers, each resolved at 250m x 250m. The
set of layers included water-cover, distance-to-water, tree-cover,
shrub/grassland-cover, and a human influence layer depicting anthropogenic
influences through villages, roads, and agriculture. A detailed description of
the different habitat layers is provided in \Cref{Hofmann.2021}. Importantly,
the water-cover and derived distance-to water layers were generated using MODIS
Terra MCD43A4 satellite imagery that was classified using a ``floodmapping''
algorithm developed by \citep{Wolski.2017} and available through the R-package
\textsf{floodmapr}. The algorithm allowed us to generate almost weekly updated
``floodmaps'', thus providing detailed information about the flood-extent at any
given point in time. In total, we generated 700 floodmaps between the years 2000
and 2019. Based on these maps, we generated a minimal and maximum flood
scenario. To create the minimum flood scenario, we averaged the 50 floodmaps
with smallest flood extent and generated a binary image using areas that were
inundated in at least 50\% of the maps. Similarly, we created an average image
for high flood using the 50 most flooded maps. The final maps are depicted in
\Cref{FloodExtent}.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_FloodExtent.png}
  \caption{Flood extent in the two scenarios considered. In the left panel, the
  flood is at an extremely low level, stretching across 3'558
  km\textsuperscript{2}, whereas in the right panel the flood is at an extremely
  high level and covers 9'397 km\textsuperscript{2}. The two maps were generated
  using 700 remote sensed MODIS MCD43A4 satellite images spanning the years 2000
  to 2019.}
  \label{FloodExtent}
  \end{center}
\end{figure}

\subsection{Dispersal Model}
Our dispersal model was based on a previously parametrized and validated
integrated step-selection function (iSSF, \citealp{Avgar.2016}) fitted to GPS
data of dispersing AWDs \citep{Hofmann.2021}. In step selection functions (SSFs,
\citealp{Fortin.2005}), observed GPS locations are converted into steps (the
straight-line traveled between two GPS recordings \citep{Turchin.1998a}) and
compared to a set of \textit{random} steps in a (mixed effects) conditional
logistic regression framework \citep{Fortin.2005, Thurfjell.2014, Muff.2020,
Fieberg.2021}. The model presented in \citep{Hofmann.2021} used dispersal data
of 16 dispersing African wild dogs from a free-ranging wild dog population in
northern Botswana. GPS data during dispersal was collected at 4-hourly intervals
and translated into steps of similar duration. Observed steps were then paired
step with 24 random steps that were generated using a uniform distribution for
turning angles (\(-\pi, +\pi\)) and step lengths from a gamma distribution
fitted to observed steps (scale \(\theta\) = 6'308 and shape \(k\) = 0.37). It
was then assumed that animals assigned to each observed and random step a
selection score of the form \citep{Fortin.2005}:

\begin{equation}
\label{EQ1}
  w(x) = exp(\beta_1 x_1 + \beta_2 x_2 + ... + \beta_n x_n)
\end{equation}

Where (\(x_1, x_2, ..., x_n\)) represent the covariate values along each of the
steps and the (\(\beta_1, \beta_2, ..., \beta_n\)) are the animal's relative
selection strengths \citealp{Avgar.2017} towards these covariates. The benefit
of \textit{integrated} SSFs over regular SSFs is that they provide a means to
render two complementary ``kernels''. A movement kernel that describes general
movement behavior of dispersing AWDs and a habitat kernel that describes
preferences of AWDs with regards to environmental conditions
\citep{Fieberg.2021}. iSSFs also allow interactions among the two kernels and
are thus suitable to render that movement behavior may change depending on
habitat conditions. A brief summary of the fitted dispersal model is provided in
Appendix xx.

\subsection{Source Areas and Emigration Zones}
We simulated dispersing AWDs originating from six distinct source areas located
in the vicinity of the Okavango delta (\Cref{StudyArea}). For areas one to six
we selected locations at the delta's periphery that remained dry in both
scenarios. At those locations, we generated circular buffers with a radius of 20
km. For source area six, we isolated a polygon covering Chief's Island, a
peninsula located at the OD's center. Besides source areas, we also generated
``zones of emigration'' that we used as checkpoints to determine if and where
simulated individuals left the delta's vicinity (\Cref{StudyArea}). We generated
emigration zones by first overlaying the OD with an elliptic buffer zone that we
dissected using a set of cutlines that originated from the ODs center and spread
according to cardinal points (\Cref{StudyArea}).

\subsection{Dispersal Simulation}
For each source area we simulated 1'000 individuals, once assuming a minimum
flood, once assuming a maximum flood. This resulted in the simulation of 6'000
individuals for each environmental scenario, hence 12'000 individuals in total.
The simulation algorithm was based on the algorithm described in
\Cref{Hofmann.2022} and works as follows. A random location within the source
area is chosen as a starting point. Originating from the starting point, a set
of 25 random steps is generated by sampling step lengths from a gamma
distribution fitted to observed steps (what is a step?) (shape = , scale = ) and
turning angles from a uniform distibution ($-\pi, +\pi$). Along each random step
the underlying spatial covariates are extracted and relevant movement metrics
are computed (i.e. log(sl), cos(ta), ta). The parametrized dispersal model is
used to predict the probability of each step for being chosen given the steps
covariate values. One of the steps is sampled based on assigned probabilities
and the location of the animal is updated. The procedure is then repated until
the desired number of steps is realized. Here, we simulated each individual for
2'000 steps, which correspons to the longest dispersal duration recorded in our
data. The original model was trained using 4-hourly steps, thus a simulated step
also resembled the movement conducted within four hours. Trajectories resulting
from such a simulation can be understood as correlated random walks that take
into account both habitat and movement preferences of dispersing individuals.

@@

\subsection{Dispersal Prospects and Connectivity}
Based on simulated dispersal trajectories in the two scenarios we quantified
dispersal sucess and connectivity using three complementary connectivity metrics
as outlined in \cite{Hofmann.2022}. First, we generated heatmaps depicting the
frequency at which different areas in the landscape were visited by simulated
dispersers. Such heatmaps serve to detect dispersal hotspots and areas of
intense use. However, they are less suitable for detecting pinchpoints and
bottlenecks that are critical in linking distinct patches. Hence, we also
computed spatially explicit betweenness scores which are useful to highlight
exactly such pinchpoints \citep{Bastille-Rousseau.2018, Bastille-Rousseau.2021}.
To compute betweenness, we overlayed the study area with a regular grid with 2.5
km x 2.5 km grid cells and determined how often simulated individuals
transitioned from one grid-cell to another. Note that in case the same
individual repeatedly realized the same cell-transition (e.g. repeatedly moved
between A-B-A-B...), we only counted a single transition to avoid emphasis on
regions where individuals moved in circles. With on the so generated
transitions, we generated a network using the centers of all grid-cells as nodes
and cell-transitions as weighted connections between the nodes. Based on this
network we computed weighted betweenness scores using the R-package
\textsf{igraph}. As a final connectivity metric and metric of dispersal success,
we calculated the number of successful dispersal events between the different
source areas as well as towards the emigration zones. We coin this type of
connectivity ``inter-patch connectivity'' as it relates to the movement between
distinct patches. Dispersal between two areas was said to be successful whenever
a trajectory leaving one area intersected with the target area. To gauge the
dispersal duration needed to move between patches (be consistent with
``patches'', ``source-areas'' etc.), we also recorded the minimum number of
steps that individuals moved before arriving at the respective patch.

\section{Results}
\subsection{Heatmaps}
Heatmaps produced from simulated dispersal trajectories reveal that the OD acts
as a major dispersal barrier during periods of high flood, but reveals viable
dispersal corridors during periods of low flood (\Cref{Heatmaps}). During
minimum flood, the area north-west of Maun appears to serve as vital dispersal
habitat. The same area is entirely avoided during maximum flood. Besides
striking differences in connectivity for the close vicinity of the delta, the
remainder of the study area shows only marginal differences in connectivity
between the two scenarios. For instance, in both scenarios the area south of the
Linyanti swamp appears as frequently visited dispersal habitat. Additional
heatmaps highlighting differences in connectivity for each source area
separately are provided in Appendix SX.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Heatmaps.png}
  \caption{Heatmaps depicting the number of simulated dispersal trajectories
  traversing each grid-cell in the study area. The left panel shows results for
  the minimum flood scenario, whereas the right panel shows results for the
  maximum flood scenario. Source areas (numbered 1-6) from which dispresers were
  released, and emigration zones (numbered 7-14) are shaded in dark gray.}
  \label{Heatmaps}
  \end{center}
\end{figure}

\subsection{Betweenness}
The betweenness maps reveal a similar pattern in that connectivity through the
OD is only pronounced during periods of low floods and vanishes entirely during
maximum flood (\Cref{Betweenness}). A set of four dispersal corridors meets on
the central peninsula (source area 5, \Cref{Betweenness}) at minimum flood but
the same corridors are absent when the flood reaches a maximum extent. Instead,
a narrow corridor runs north west of Maun, connecting source areas one and two.
Again, the remainder of the study area is only marginally affected by flooding
patterns. Additional betweenness maps highlighting differences in connectivity
for each source area separately are provided in Appendix SX.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Betweenness.png}
  \caption{Betweenness scores highlighting potential dispersal corridors for
  both the minimum and maximum flood scenario. Areas with high betweenness
  scores (bright yellow) are used by many simulated individuals to move into
  adjacent regions and can thus be understood as critical pinch-points. Source
  areas (numbered 1-6) from which dispresers were released, and emigration zones
  (numbered 7-14) are shaded in light gray.}
  \label{Betweenness}
  \end{center}
\end{figure}

\subsection{Inter-Patch Connectivity}
Our analysis of interpatch connectivity demonstrates notable differences in
dispersal prospects and duration depending on the extent of the flood
(\Cref{IPCTable}). While $\input{99_NumberReachingOthersMinFlood.tex}$ simulated
dispersers reach another source area during minimum extent, only
$\input{99_NumberReachingOthersMaxFlood.tex}$ do so during maximum extent. The
differences are particularly pronounced for individuals dispersing from or into
the source area located at the OD's center (\Cref{Interpatch}). While the area
is reached by $\input{99_NumberReachingArea6MinFlood.tex}$ simulated individuals
during minimum flood, only $\input{99_NumberReachingArea6MaxFlood.tex}$
dispersers arrive there during maximum flood. Furthermore, the dispersal
duration into source area six from any other source area increases from
$\input{99_StepsToReachingArea6MinFlood.tex}$ to
$\input{99_StepsToReachingArea6MaxFlood.tex}$. Across all simulations, the
average dispersal duration before reaching another source area increases from xx
to xx from the minimum to the maximum flood scenario. Nevertheless, connectivity
into some areas increases during maximum flooding. Emigration increased slightly
from xx to xx

For instance, while dispersers .. Additional maps highlighting differences in
inter-patch connectivty for each source area separately are provided in Appendix
SX.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_IPCTable.png}
  \caption{Dispersal frequency (a) and duration (b) (in steps) between source
  areas and emigration zones during minimum and maximum flood.}
  \label{IPCTable}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Interpatch.png}
  \caption{Inter-patch connectivity into and from source area number six. This
  patch experienced the most drastic reduction in connectivity in result to a
  high flood level. During minimum flood there is ample connectivity from and
  into source areas 1-3. During maximum flood, connectivity is lower and mainly
  limited to source area 2.}
  \label{Interpatch}
  \end{center}
\end{figure}

% \input{IPC_Frequency.tex}
% \input{IPC_Duration.tex}
% \input{IPC.tex}

@@

\cite{Murray-Hudson.2006} predicted that increased temperatures, additional human
abstractions, and reduced river flows might lead to a ``Delta dying'' and that
the impact of climate will be much more pronounced than the impact of
anthropogenic water use.

@@

We studied a population of African wild dogs that resides in a natural
environment with little human influence. This is only representative for a small
share of the extant wild dog populations, as most individuals reside in areas
that are prone to substantial edge effects. For these populations, the benefit
of dispersal is disputed, as dispersal under high anthropogenic mortality may
lead to a net-loss of individuals \citep{Leigh.2012}.

We assessed the implications of environmental change on the dispersal prospects,
yet we did not consider how changing conditions alter dispersal propensity.

@@

\newpage
\begingroup
\singlespacing
\bibliography{Literature}
\endgroup
